---
title: "ClockSHAP Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clockshap-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(ClockSHAP)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction

ClockSHAP is a method for interpreting *biological age deviation* produced by
linear aging clocks.

Rather than explaining which features increase or decrease a predicted age,
ClockSHAP explains **why a sample appears biologically older or younger than
expected for its chronological age**, relative to a reference aging trajectory.

The method provides an exact, closed-form decomposition for linear models.

---

## Conceptual overview

Consider a linear aging clock trained on reference data:

\[
\hat{y} = \alpha + \sum_k \beta_k \tilde{X}_k
\]

ClockSHAP defines deviation as:

\[
\Delta = \hat{y} - E[\hat{y} \mid \text{Age}]
\]

and decomposes this deviation into per-feature contributions:

\[
\phi_k = \beta_k \left( \tilde{X}_k - E[\tilde{X}_k \mid \text{Age}] \right)
\]


such that:

\[
\sum_k \phi_k = \Delta
\]

where \(\tilde{X}_k\) denotes features standardized using the clock’s training mean and standard deviation.

This decomposition is exact for linear clocks.

---

## A minimal example

This example uses synthetic data to demonstrate the full ClockSHAP workflow.

```{r}
set.seed(1)

features <- matrix(rnorm(20), nrow = 5, ncol = 4)
colnames(features) <- paste0("F", 1:4)

age <- c(40, 50, 60, 70, 80)

clock <- linear_clock(
  alpha = 10,
  beta  = setNames(runif(4), colnames(features)),
  mu    = setNames(rep(0, 4), colnames(features)),
  sigma = setNames(rep(1, 4), colnames(features))
)

reference <- fit_reference_profile(
  features = as.data.frame(features),
  age = age
)

cs <- clockshap(features, age, clock, reference)
cs
```

## Inspecting results
ClockSHAP returns a structured object with accessor functions.

```{r}
clockshap_deviation(cs)
clockshap_phi(cs)
```

By construction, per-feature contributions sum to the total deviation for each
sample.

```{r}
all.equal(rowSums(clockshap_phi(cs)), clockshap_deviation(cs), tolerance = 1e-10)
```

## Waterfall visualization

```{r}
plot_clockshap_waterfall(cs, sample = 1)
```

## Scope and assumptions
ClockSHAP is defined for linear aging clocks and requires:

- linear feature contributions
- a reference age-feature relationship
- consistent feature definitions between clock and reference

The method does not train clocks or impose assumptions about the source of
reference data. Any dataset may be used to define a reference aging trajectory.

