---
title: "Introduction to ClockSHAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ClockSHAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  out.width = "100%"
)
library(ClockSHAP)
```

## Overview

ClockSHAP interprets the output of a linear aging clock. Instead of asking which
features push a predicted age up or down in absolute terms, it answers:

> **Why is this sample predicted to be biologically older or younger than expected for its chronological age?**

For a linear clock it provides an exact, closed-form decomposition of that gap
into per-feature contributions. This vignette walks through the full workflow on
the package's bundled synthetic dataset. The data, the core objects, the
decomposition, the accessors, and how the result relates to the more familiar
age-acceleration metrics.

## The method in brief

A linear clock predicts age, $\hat{y}$, from features. ClockSHAP defines the
**age-matched deviation** as the gap between that prediction and the age-matched
expectation under a reference:

$$\Delta_i = \hat{y}_i - \mathrm{E}[\hat{y} \mid \mathrm{Age}_i]$$

The expectation comes from an **age-conditioned reference profile** — a
per-feature linear age trend fit on a reference cohort:

$$\mathrm{E}[X_k \mid \mathrm{Age}] = \gamma_{0,k} + \gamma_{1,k}\,\mathrm{Age}$$

Each feature's contribution is its clock weight $w_k$ times its distance from
that age-expected value, and for a linear clock the contributions reconstruct the
deviation exactly:

$$\phi_{i,k} = w_k\big(x_{i,k} - \mathrm{E}[X_k \mid \mathrm{Age}_i]\big), \qquad \sum_k \phi_{i,k} = \Delta_i$$

In practice features are standardized to the clock's training distribution before
the decomposition is computed.

## The example data

The package ships a small, fully synthetic dataset, `clockshap_example`, so every
example runs without external data. It is a list with three pieces:

```{r data-structure}
str(clockshap_example, max.level = 1)
```

`features` is a 120 × 8 matrix of synthetic "gene" values (rows `S001`–`S120`),
each gene carrying a genuine linear age trend plus noise; `age` is the
chronological age; and `clock` is a fitted clock object. Two samples are planted
deviators: **`S001`** reads older than expected for its age and **`S002`** reads
younger, both driven by a shared senescence / inflammaging axis (`CDKN2A`,
`GDF15`, `IL6`). The data carry no real biological signal and the gene symbols are
only there to make the output legible.

## The building blocks: clock and reference

ClockSHAP works with two specification objects.

A **`linear_clock`** holds the intercept, the per-feature weights, and the
training mean and standard deviation used to standardize features. You construct
one with `linear_clock()`; here we inspect the bundled clock:

```{r clock}
clock <- clockshap_example$clock
class(clock)
round(clock$beta, 2)   # per-gene weights learned by the clock
```

A **`reference_profile`** holds, for each feature, the intercept and slope of its
age trend ($\gamma_0$, $\gamma_1$). You can build one directly with
`reference_profile(gamma0, gamma1)`, but usually you fit it from a cohort with
`fit_reference_profile()`, which regresses each feature on age:

```{r reference}
ref <- fit_reference_profile(
  features = as.data.frame(clockshap_example$features),
  age      = clockshap_example$age
)
class(ref)
round(ref$gamma1, 4)   # per-gene age slopes recovered from the cohort
```

The non-zero slopes confirm the features genuinely move with age, which is what
makes the age-conditioned expectation meaningful.

## Running ClockSHAP

`clockshap()` ties the pieces together: features, ages, a clock, and a reference.
It standardizes the features with the clock's parameters, builds the age-matched
expectation from the reference, and returns the decomposition.

```{r run}
cs <- clockshap(
  features  = clockshap_example$features,
  age       = clockshap_example$age,
  clock     = clockshap_example$clock,
  reference = ref
)

cs                # print method: dimensions + deviation range
summary(cs)$deviation
```

`clockshap()` also enforces its core invariant internally — it will error rather
than return a result where the contributions fail to sum to the deviation.

## The output: structure and accessors

A `clockshap` object is a list with five elements, each retrievable either by
`$` or with an accessor function:

| field | accessor | meaning |
|---|---|---|
| `predicted` | `clockshap_predicted()` | clock prediction $\hat{y}$ |
| `expected`  | `clockshap_expected()`  | age-matched expectation $\mathrm{E}[\hat{y}\mid\mathrm{Age}]$ |
| `deviation` | `clockshap_deviation()` | $\Delta = \text{predicted} - \text{expected}$ |
| `phi`       | `clockshap_phi()`       | per-feature contribution matrix (samples × features) |
| `age`       | —                       | the input chronological ages |

```{r results}
head(data.frame(
  age       = cs$age,
  predicted = round(clockshap_predicted(cs), 1),
  expected  = round(clockshap_expected(cs), 1),
  deviation = round(clockshap_deviation(cs), 1)
))
```

The contributions in `clockshap_phi()` sum, row by row, exactly to the deviation:

```{r additivity}
all.equal(rowSums(clockshap_phi(cs)), clockshap_deviation(cs), tolerance = 1e-10)
```

## Cohort view

Plotting each sample's predicted age against its chronological age reproduces the
cohort-level picture. The black line is the age-matched expectation; a sample's
vertical distance from it is its deviation. The two planted deviators sit clearly
off the line.

```{r cohort, fig.alt = "Predicted vs chronological age across the synthetic cohort, with two deviators highlighted"}
library(ggplot2)

cohort <- data.frame(
  age       = cs$age,
  predicted = clockshap_predicted(cs),
  role      = ifelse(rownames(clockshap_example$features) == "S001", "S001 (older)",
               ifelse(rownames(clockshap_example$features) == "S002", "S002 (younger)",
                      "cohort"))
)

ggplot(cohort, aes(age, predicted)) +
  geom_point(data = subset(cohort, role == "cohort"),
             colour = "grey70", alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.6) +
  geom_point(data = subset(cohort, role != "cohort"),
             aes(colour = role), size = 3) +
  scale_colour_manual(values = c("S001 (older)" = "#B2182B",
                                 "S002 (younger)" = "#2166AC")) +
  labs(x = "Chronological age (years)", y = "Predicted age (years)", colour = NULL) +
  theme_minimal()
```

## Age-matched deviation vs. AAA and RAA

It helps to see how the deviation relates to two metrics common in the
aging-clock literature:

- **Absolute age acceleration (AAA):** $\hat{y} - \mathrm{Age}$.
- **Relative age acceleration (RAA):** the residuals of `lm(predicted ~ age)`.

```{r metrics}
age  <- cs$age
pred <- clockshap_predicted(cs)

metrics <- data.frame(
  AAA       = pred - age,
  RAA       = as.numeric(residuals(lm(pred ~ age))),
  deviation = clockshap_deviation(cs)
)
```

Because this clock has a slope below 1, AAA is **age-biased** — systematically
positive for younger samples and negative for older ones — while RAA and the
deviation are not:

```{r metrics-cor}
round(sapply(metrics, function(z) cor(z, age)), 3)
```

And here the reference was fit on the same cohort we are analyzing. In that
self-reference case the age-matched deviation is **exactly** relative age
acceleration:

```{r metrics-equal}
all.equal(metrics$deviation, metrics$RAA, tolerance = 1e-9)
```

This is not a coincidence. For a linear clock, passing per-feature age trends
through the clock reproduces the OLS fit of predicted age on age, so
`predicted - expected` equals the regression residual. ClockSHAP becomes distinct
from RAA — and starts carrying new biological signal — only when the reference is
a *different*, biologically meaningful cohort (e.g., a tissue-of-origin
reference). What it always adds, even in the self-reference case, is the
**per-feature attribution** of that deviation, which neither AAA nor RAA provides.

## Per-sample decomposition

`plot_clockshap_waterfall()` shows, for a single sample, how each feature carries
the prediction from the age-matched expectation to the predicted age. The bars
sum to the deviation. Compare the two planted deviators:

```{r waterfall-pos, fig.height = 3.4, fig.alt = "Waterfall for S001, an older-than-expected sample"}
plot_clockshap_waterfall(cs, sample = "S001", top_n = 6)
```

```{r waterfall-neg, fig.height = 3.4, fig.alt = "Waterfall for S002, a younger-than-expected sample"}
plot_clockshap_waterfall(cs, sample = "S002", top_n = 6)
```

`S001` reads older because its senescence / inflammaging genes sit above their
age-expected levels; `S002` reads younger for the mirror-image reason. In both,
the labeled contributions add up to the age-matched deviation.

## Scope and assumptions

ClockSHAP is defined for linear aging clocks and requires:

- linear, additive feature contributions (a `linear_clock`),
- an age–feature reference relationship (a `reference_profile`),
- consistent feature definitions between the clock and the reference.

It does not train clocks, and it does not assume a particular source of reference
data — any cohort can define the reference aging trajectory. Because the choice of
reference determines what "expected for age" means, deviations should always be
interpreted relative to it.
